### 2015-04-14 src/paper4/paper4.tex
References

\small {\bf Abstract}
It has been shown that the compensatory eye movements, that minimize retinal slip during self-motion, also serve as a cue for translation perception. However, to provide an adequate translation estimate, the brain must internally scale these ensuing eye movements by fixation distance. Using a 2-AFC approach, we investigated whether the brain applies this scaling.  Participants ($n = 8$) were translated sideways in the absence of full-field optic flow but with gaze maintained on either a nearby or faraway target, that was either fixed in the world (world-fixed) or moved along with the body (body-fixed). Results show that translations were perceived shorter with gaze on nearby than faraway world-fixed targets, indicating that eye movements are not properly scaled in translation perception. Translation perception was not affected by the depth of body-fixed targets. Taken together, our results suggest that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

@@

When the eyes track world-centered objects, their angular displacement is directly related to the size of the motion of the observer \cite{schwarz1989, paige1998, mchenry2000, medendorp2002}. Because  the majority of fixations are on world-stationary  objects, we recently proposed that these tracking eye movements could also be used as a self-motion cue, in addition to optic flow and vestibular signals. To test this hypothesis, we compared self-motion perception in the absence of full-field optic flow during passively induced whole-body translations \cite{clemens2015a}. Our results showed that self-motion is underestimated during body-centered fixations (in which the eyes remain stationary in their orbits) compared to fixations on world-stationary objects (in which the eyes must move to maintain fixation).

Geometrically, eye movements that keep fixation on a world-centered target during lateral whole body translation (i.e. the linear vestibulo-ocular reflex; LVOR), must scale with fixation depth \cite{angelaki2004}. When fixating body-centered targets these eye movements must be suppressed irrespective of fixation distance \cite{angelaki2004}. Conversely, when fixating world-centered targets, the brain must internally scale the ensuing eye movement by fixation distance to serve as an adequate self-motion cue. Because we did not manipulate fixation distance in our previous study, we could not dissociate whether eye movements are used as a rudimentary cue for self-motion (i.e. without taking fixation depth into account), or are properly scaled in the mechanisms for self-motion perception.

In the present study, we investigate how fixation distance influences perception of self-motion during passive side-to-side translations. Using a psychophysical approach, participants had to indicate whether the second body displacement of two one-second translation intervals was smaller or longer than the first. We show that translation amplitude is perceived smaller when fixating a far compared to a nearby world-centered target, indicating that eye movements are not properly scaled in self-motion perception. Together with the observation that self-motion perception is not affected by the depth of a body-centered fixation target, we conclude that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

@@

Participants were seated on a motorized linear sled with their body and head restrained such that the inter-aural axis aligned with the motion axis. The sled laterally translated participants following a minimum jerk profile of fixed duration (1 \si{\second}) and amplitudes ranging from 1 to 27 \si{\centi\metre}. Auditory cues were suppressed using white noise presented through in-ear head-phones. Experiments were conducted in complete darkness except for visual fixation points, projected by body- or world-fixed laser pointers on a black bar, either 50 or 200 \si{\centi\metre} in front of the participant and at eye level. Eye movements were recorded at 500Hz using an EyeLink II system (SR Research, Kanata, Canada). Eye position was calibrated before every session using 11 evenly spaced calibration points ranging from -22 to 22 degrees.

@@

We used a two-alternative forced choice (2-AFC) task to study the influence of fixation depth on the perception of linear translation. We tested two fixation depths: near (50 \si{\centi\metre}) and far (200 \si{\centi\metre}) and two different eye fixation conditions: world-, and body-fixed fixation. A trial contained two sequential motion intervals of equal duration (1 \si{\second}), with the motion in the same direction (either leftward or rightward).

@@

The timing of a single trial is shown in \panelref{p4:fig1}{A}. Every trial started with the onset of a central fixation point (i.e. aligned between the eyes), at a depth of 50 or 200\si{\centi\metre}, for 0.5 \si{\second}. Subsequently, the first 1\si{\second} motion interval commenced. During this translation, the fixation point either remained world stationary (world condition), or moved along with the participant (body condition).  Subsequently a 1.75 \si{\second} delay followed in which the sled was stationary and no  fixation light was shown. Next, a central fixation point reappeared at the other depth (50 or 200\si{\centi\metre}) than was used in the first interval. After 0.5 \si{\second}, the second 1 \si{\second} translation interval started, with the same fixation condition as in the first interval. After this second interval, the participant responded whether the second displacement was perceived longer or shorter than the first by moving a 1-dimensional joystick away from (longer) or towards (shorter) the body. Top-view illustrations of example body-centered and world-centered trials are shown in \panelref{p4:fig1}{B} and \hyperref[p4:fig1]{C} respectively.

@@

For each combination of the two main conditions and the two reference/probe orders we computed the probability $P(x)$ of probe translation $x$ being  judged as longer than the reference translation. To summarize these data, we fit cumulative Gaussian functions to these probabilities, resulting in a total of four Gaussian functions per participant (see \tabref{p4:tab1}):

@@

We fit \eqnref{p4:eq5} to the data using linear regression, finding one weight for each of the two fixation depths (that is, $\alpha_{50}$ and $\alpha_{200}$) that minimizes the sum of squared errors ($\Sigma \epsilon^2$). Because these parameters can, in theory, contain both a depth-dependent and a depth-independent scaling, we compute their ratio, $\frac{\alpha_{200}}{\alpha_{50}}$, to remove any depth independent components. In case of perfect compensation, the expected ratio is $\frac{200}{50} = 4$, while in case of depth-independent scaling it is 1.

@@

    \caption{\panel{A} Actual (solid lines) and ideal (dashed lines) eye movement traces of one participant in the body-fixed (brown and orange) and world-fixed conditions (purple and pink). Gaze was directed at a near (brown and purple) or far (orange and pink) target. All traces shown are for 10m reference displacements. \panel{B} Normalized eye position for each participant (\textpm 95\% confidence interval) at the end of the translation interval for the near and far world fixed targets (purple and pink respectively) as well as the near and far body fixed fixation targets (brown and orange respectively). In addition, the average \textpm SE across all participants is shown. Zero indicates the eyes remained stationary relative to the body, and one indicates the eyes followed the near world fixed target perfectly.}
    \label{p4:fig2}
\end{figure}

@@

We normalized the eye movement data by taking the average ratio between the measured eye excursion and the geometrically required eye displacement were the target world stationary. \panelref{p4:fig2}{B} shows that these normalized eye displacements are about zero during body fixed fixations (brown and orange) and close to one in the world fixed fixations (purple and pink data points), for all participants. The question is whether these eye movements are inversely scaled by fixation depth in order to interpret them as linear self-motion cues.

@@

	\caption{\panel{A} and \panel{B} Psychometric curves (colored lines) and associated binned data (circles) for one participant. Circle size represents the amount of trials within the bin. Psychometric curves before collapsing across reference order are shown as gray lines. \panel{A} World-fixed condition (purple) while fixation was either near (dark) or far (light). \panel{B} Body-fixed condition (brown) while fixation was either near (dark) or far (light).
	\panel{C} and \panel{D} Individual and average points of subjective equality (PSEs). Color scheme matches panels A and B.
	}
	\label{p4:fig3}
\end{figure}

\figref{p4:fig3} illustrates psychophysical data on self-motion perception of a single participant for the two-fixation depths in both the world (\panelref{p4:fig3}{A}) and body condition (\panelref{p4:fig3}{B}). Lighter and darker colors indicate which fixation depth was the reference translation (see figure legend). The influence of fixation depth is characterized by a shift of the psychometric functions relative to the 10 \si{\centi\metre} reference translation (i.e. the PSE). For example, the rightward shift of the pink curve in \figref{p4:fig3}A, representing the world-condition, means that with a far target a longer translation (\about15 \si{\centi\metre}) was required to be perceived equivalent to a 10 \si{\centi\metre} reference translation with nearby fixation. Likewise, the leftward shift of the purple curve indicates that a shorter translation with near fixation (\about6 \si{\centi\metre}) is required to be perceived the same as the 10 \si{\centi\metre} reference translation with far fixation. Together, these opposite biases suggest that translations are perceived shorter for fixations further away. For the body condition, no shift of the psychometric curves is visible (\panelref{p4:fig3}{B}), indicating that fixation depth (i.e. near versus far) has no effect in absence of eye movements.

@@

	\caption{Eye movement based prediction for the PSE plotted against the actual PSE. A data point (symbol) is shown for each participant (symbol shape) and condition (symbol color) pair, following the same color scheme as in \figref{p4:fig3}. The identity line, corresponding to a perfect prediction is shown (solid line) as well as the best fit line (dashed).}
	\label{p4:fig4}
\end{figure}

@@

% Model results
To quantify the relative depth-dependent scaling of eye movements for nearby and far away fixation targets, we fitted a straightforward linear model to the perceptual responses based on the oculomotor behavior across four conditions. While two participants show partial scaling, the other six participants did not show any sign of scaling. Thus, we conclude that while oculomotor signals provide a robust cue to translation perception, they are not properly scaled by fixation depth.

@@

    \caption{Predicted versus actual PSEs for Clemens \protect\citeyear{clemens2015a} using parameter $\alpha_{50}$ from the present paper. A data point (symbol) is shown for each participant (symbol shape) and condition (symbol color) pair, following the same color scheme as in \protect\figref{p3:fig2}: Body-world comparison; body reference, dark red; world reference, light red. World-free comparison; world reference, light green; free reference, dark green. Body-free comparison; body reference, dark blue; free reference, light blue.}
    \label{p4:fig6}
\end{figure}

In our previous experiment we compared translation perception with body-fixed versus world-fixed fixations at near depth (50 \si{\centi\metre}) only \cite{clemens2015a}. \figref{p4:fig6} shows how well our $\alpha_{50}$ parameter explains the data in our previous paper. The positive correlation between the actual PSEs in the previous study and those predicted using the model of the present paper ($\rho = 0.60$, $p = 0.06$) adds confidence to the parameter values presented here. The average difference between the values found here and those reported previously (see \tabref{p4:tab2}) is 12 \textpm 8 percent-points, which is relatively small given the indpendent measurements.

>>>>

\small {\bf Abstract}
It has been shown that the compensatory eye movements, that minimise retinal slip during self-motion, also serve as a cue for translation perception. However, to provide an adequate translation estimate, the brain must internally scale these ensuing eye movements by fixation distance. Using a 2AFC approach, we investigated whether the brain applies this scaling.  Participants ($n = 8$) were translated sideways in the absence of full-field optic flow but with gaze maintained on either a nearby or faraway target, that was either fixed in the world (world-fixed) or moved along with the body (body-fixed). Results show that translations were perceived shorter with gaze on nearby than faraway world-fixed targets, indicating that eye movements are not properly scaled in translation perception. Translation perception was not affected by the depth of body-fixed targets. Taken together, our results suggest that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

@@

When the eyes track world-centred objects, their angular displacement is directly related to the size of the motion of the observer \cite{schwarz1989, paige1998, mchenry2000, medendorp2002}. Because  the majority of fixations are on world-stationary  objects, we recently proposed that these tracking eye movements could also be used as a self-motion cue, in addition to optic flow and vestibular signals. To test this hypothesis, we compared self-motion perception in the absence of full-field optic flow during passively induced whole-body translations \cite{clemens2015a}. Our results showed that self-motion is underestimated during body-centred fixations (in which the eyes remain stationary in their orbits) compared to fixations on world-stationary objects (in which the eyes must move to maintain fixation).

Geometrically, eye movements that keep fixation on a world-centred target during lateral whole body translation (i.e. the linear vestibulo-ocular reflex; LVOR), must scale with fixation depth \cite{angelaki2004}. When fixating body-centred targets these eye movements must be suppressed irrespective of fixation distance \cite{angelaki2004}. Conversely, when fixating world-centred targets, the brain must internally scale the ensuing eye movement by fixation distance to serve as an adequate self-motion cue. Because we did not manipulate fixation distance in our previous study, we could not dissociate whether eye movements are used as a rudimentary cue for self-motion (i.e. without taking fixation depth into account), or are properly scaled in the mechanisms for self-motion perception.

In the present study, we investigate how fixation distance influences perception of self-motion during passive side-to-side translations. Using a psychophysical approach, participants had to indicate whether the second body displacement of two one-second translation intervals was smaller or longer than the first. We show that translation amplitude is perceived smaller when fixating a far compared to a nearby world-centred target, indicating that eye movements are not properly scaled in self-motion perception. Together with the observation that self-motion perception is not affected by the depth of a body-centred fixation target, we conclude that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

@@

Participants were seated on a motorised linear sled with their body and head restrained such that the inter-aural axis aligned with the motion axis. The sled laterally translated participants following a minimum jerk profile of fixed duration (1 \si{\second}) and amplitudes ranging from 1 to 27 \si{\centi\metre}. Auditory cues were suppressed using white noise presented through in-ear head-phones. Experiments were conducted in complete darkness except for visual fixation points, projected by body- or world-fixed laser pointers on a black bar, either 50 or 200 \si{\centi\metre} in front of the participant and at eye level. Eye movements were recorded at 500Hz using an EyeLink II system (SR Research, Kanata, Canada). Eye position was calibrated before every session using 11 evenly spaced calibration points ranging from -22 to 22 degrees.

@@

We used a two-alternative forced choice (2AFC) task to study the influence of fixation depth on the perception of linear translation. We tested two fixation depths: near (50 \si{\centi\metre}) and far (200 \si{\centi\metre}) and two different eye fixation conditions: world-, and body-fixed fixation. A trial contained two sequential motion intervals of equal duration (1 \si{\second}), with the motion in the same direction (either leftward or rightward).

@@

The timing of a single trial is shown in \panelref{p4:fig1}{A}. Every trial started with the onset of a central fixation point (i.e. aligned between the eyes), at a depth of 50 or 200\si{\centi\metre}, for 0.5 \si{\second}. Subsequently, the first 1\si{\second} motion interval commenced. During this translation, the fixation point either remained world stationary (world condition), or moved along with the participant (body condition).  Subsequently a 1.75 \si{\second} delay followed in which the sled was stationary and no  fixation light was shown. Next, a central fixation point reappeared at the other depth (50 or 200\si{\centi\metre}) than was used in the first interval. After 0.5 \si{\second}, the second 1 \si{\second} translation interval started, with the same fixation condition as in the first interval. After this second interval, the participant responded whether the second displacement was perceived longer or shorter than the first by moving a 1-dimensional joystick away from (longer) or towards (shorter) the body. Top-view illustrations of example body-centred and world-centred trials are shown in \panelref{p4:fig1}{B} and \hyperref[p4:fig1]{C} respectively.

@@

For each combination of the two main conditions and the two reference/probe orders we computed the probability $P(x)$ of probe translation $x$ being  judged as longer than the reference translation. To summarise these data, we fit cumulative Gaussian functions to these probabilities, resulting in a total of four Gaussian functions per participant (see \tabref{p4:tab1}):

@@

We fit \eqnref{p4:eq5} to the data using linear regression, finding one weight for each of the two fixation depths (that is, $\alpha_{50}$ and $\alpha_{200}$) that minimises the sum of squared errors ($\Sigma \epsilon^2$). Because these parameters can, in theory, contain both a depth-dependent and a depth-independent scaling, we compute their ratio, $\alpha_{200}/\alpha_{50}$, to remove any depth independent components. In case of perfect compensation, the expected ratio is $200/50 = 4$, while in case of depth-independent scaling it is 1.

@@

    \caption{\panel{A} Actual (solid lines) and ideal (dashed lines) eye movement traces of one participant in the body-fixed (brown and orange) and world-fixed conditions (purple and pink). Gaze was directed at a near (brown and purple) or far (orange and pink) target. All traces shown are for 10m reference displacements. \panel{B} Normalised eye position for each participant (\textpm 95\% confidence interval) at the end of the translation interval for the near and far world fixed targets (purple and pink respectively) as well as the near and far body fixed fixation targets (brown and orange respectively). In addition, the average \textpm SE across all participants is shown. Zero indicates the eyes remained stationary relative to the body, and one indicates the eyes followed the near world fixed target perfectly.}
    \label{p4:fig2}
\end{figure}

@@

We normalised the eye movement data by taking the average ratio between the measured eye excursion and the geometrically required eye displacement were the target world stationary. \panelref{p4:fig2}{B} shows that these normalised eye displacements are about zero during body fixed fixations (brown and orange) and close to one in the world fixed fixations (purple and pink data points), for all participants. The question is whether these eye movements are inversely scaled by fixation depth in order to interpret them as linear self-motion cues.

@@

	\caption{\panel{A} and \panel{B} Psychometric curves (coloured lines) and associated binned data (circles) for one participant. Circle size represents the amount of trials within the bin. Psychometric curves before collapsing across reference order are shown as gray lines. \panel{A} World-fixed condition (purple) while fixation was either near (dark) or far (light). \panel{B} Body-fixed condition (brown) while fixation was either near (dark) or far (light).
	\panel{C} and \panel{D} Individual and average points of subjective equality (PSEs). Colour scheme matches panels A and B.
	}
	\label{p4:fig3}
\end{figure}

\figref{p4:fig3} illustrates psychophysical data on self-motion perception of a single participant for the two-fixation depths in both the world (\panelref{p4:fig3}{A}) and body condition (\panelref{p4:fig3}{B}). Lighter and darker colours indicate which fixation depth was the reference translation (see figure legend). The influence of fixation depth is characterised by a shift of the psychometric functions relative to the 10 \si{\centi\metre} reference translation (i.e. the PSE). For example, the rightward shift of the pink curve in \panelref{p4:fig3}{A}, representing the world-condition, means that with a far target a longer translation (\about15 \si{\centi\metre}) was required to be perceived equivalent to a 10 \si{\centi\metre} reference translation with nearby fixation. Likewise, the leftward shift of the purple curve indicates that a shorter translation with near fixation (\about6 \si{\centi\metre}) is required to be perceived the same as the 10 \si{\centi\metre} reference translation with far fixation. Together, these opposite biases suggest that translations are perceived shorter for fixations further away. For the body condition, no shift of the psychometric curves is visible (\panelref{p4:fig3}{B}), indicating that fixation depth (i.e. near versus far) has no effect in absence of eye movements.

@@

	\caption{Eye movement based prediction for the PSE plotted against the actual PSE. A data point (symbol) is shown for each participant (symbol shape) and condition (symbol colour) pair, following the same colour scheme as in \figref{p4:fig3}. The identity line, corresponding to a perfect prediction is shown (solid line) as well as the best fit line (dashed).}
	\label{p4:fig4}
\end{figure}

@@

% Model results
To quantify the relative depth-dependent scaling of eye movements for nearby and far away fixation targets, we fitted a straightforward linear model to the perceptual responses based on the oculomotor behaviour across four conditions. While two participants show partial scaling, the other six participants did not show any sign of scaling. Thus, we conclude that while oculomotor signals provide a robust cue to translation perception, they are not properly scaled by fixation depth.

@@

    \caption{Predicted versus actual PSEs for Clemens \protect\citeyear{clemens2015a} using parameter $\alpha_{50}$ from the present paper. A data point (symbol) is shown for each participant (symbol shape) and condition (symbol colour) pair, following the same colour scheme as in \protect\figref{p3:fig2}: Body-world comparison; body reference, dark red; world reference, light red. World-free comparison; world reference, light green; free reference, dark green. Body-free comparison; body reference, dark blue; free reference, light blue.}
    \label{p4:fig6}
\end{figure}

In our previous experiment we compared translation perception with body-fixed versus world-fixed fixations at near depth (50 \si{\centi\metre}) only \cite{clemens2015a}. \figref{p4:fig6} shows how well our $\alpha_{50}$ parameter explains the data in our previous paper. The positive correlation between the actual PSEs in the previous study and those predicted using the model of the present paper ($\rho = 0.60$, $p = 0.06$) adds confidence to the parameter values presented here. The average difference between the values found here and those reported previously (see \tabref{p4:tab2}) is 12 \textpm 8 percent-points, which is relatively small given the independent measurements.
